---
title: "Class 9: Candy Mini-Project"
author: "Jacob Hizon A17776679"
format: pdf
toc: TRUE
---



## Importing candy data 

```{r}
candy_file <- "candy-data.csv"
```

```{r}
candy = read.csv(candy_file, row.names=1)
head(candy)
```

> Q1. How many different candy types are in this dataset?

There are 85 different types of candy in this dataset

```{r}
nrow(candy)
```

> Q2. How many fruity candy types are in the dataset?


There are 38 types of fruity candy. 
```{r}
sum(candy$fruity)
```

```{r}
candy["Twix", ]$winpercent
```

```{r}
library(dplyr)

candy |> 
  filter(row.names(candy)=="Twix") |> 
  select(winpercent)
```

> Q3. What is your favorite candy (other than Twix) in the dataset and what is it’s winpercent value?


```{r}
candy["100 Grand", ]$winpercent
```
66.98% 


> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

```{r}
library("skimr")
skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Yes, the winpercent.  


> Q7. What do you think a zero and one represent for the candy$chocolate column?

It represents zero not a chocolate, and a 1 represents true or is a chocolate candy. 


## Exploratory analysis

> Q8. Plot a histogram of winpercent values using both base R an ggplot2

```{r}
hist(candy$winpercent)
```


```{r}
library(ggplot2)
ggplot(candy, aes(winpercent)) + 
  geom_histogram()
```

> Q9. Is the distribution of winpercent values symmetrical?

No, the distribution is not symetrical.

> Q10.  Is the center of the distribution above or below 50%?

```{r}
median(candy$winpercent)
```
Below 50%.

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
chocwin <- candy$winpercent[as.logical(candy$chocolate)]
fruitwin <- candy$winpercent[as.logical(candy$fruity)]

t.test(chocwin, fruitwin)
```
> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
mean(chocwin)
mean(fruitwin)
```
Chocolate candies have a higher average winpercent than fruity candies.

>Q12. Is this difference statistically significant?

Yes this difference is statistically significant because the p-value is extremely low. Therefore, we can say that the difference in winpercent between chocolate and fruity candies is not due to random chance.


## Overall candy rankings


```{r}
order(candy$winpercent)
```


```{r}
candy %>%
  arrange(winpercent) %>%
  head(5)

```
> Q13. What are the five least liked candy types in this set?

Nik L Nip, Boston Baked Beans, Chicklets, Super Bubble, Jawbusters 

> Q14. What are the top 5 all time favorite candy types out of this set?

Reese's Peanut Butter cup, Reese's Minatures, Twix, Kit kat, snickers. 
```{r}
candy %>%
  arrange(desc(winpercent)) %>%
  head(5)
```


```{r}
head(candy[order(candy$winpercent),], n=5)
```
> Q15. Make a first barplot of candy ranking based on winpercent values.


```{r}
library(ggplot2)
ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```
> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col()
```

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

> Q17. What is the worst ranked chocolate candy?

Nik L Lip 

> Q18. What is the best ranked fruity candy?

Starburst 


## Taking a look at pricepercent


```{r}
library(ggrepel)

# How about a plot of win vs price
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```
> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's minatures 

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

The tope 5 most expensive candy and the least popular are Nik L Nip, Nestle smarties, Ring pop, Hershey's Krackel, and hershey's milk chocolate respectively. 

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

> Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().




```{r}
# Make a lollipop chart of pricepercent
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```



## Exploring the correlation structure

```{r}
library(corrplot)
```
```{r}
cij <- cor(candy)
corrplot(cij)
```
> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and fruity.
> Q23. Similarly, what two variables are most positively correlated?

Chocolate and win percent. 

## Principal Component Analysis


```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```

```{r}
pca <- prcomp(candy, scale = TRUE)

my_data <- cbind(candy, pca$x[, 1:3])
```

```{r}
p <- ggplot(my_data) + 
  aes(x = PC1, 
      y = PC2, 
      size = winpercent / 100,  
      text = rownames(my_data),
      label = rownames(my_data)) +
  geom_point(col = my_cols)

p
```


```{r}
ggplot(pca$x) + 
  aes(PC1, PC2, label = row.names(pca$x)) +
  geom_point(col = my_cols) +
  geom_text_repel(max.overalps = 5, size = 3.3, col = my_cols)

```
> Q24. Complete the code to generate the loadings plot above. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? Where did you see this relationship highlighted previously?

Fruity, plurbis, and hard are picked up strongly by PC1 in the positive direction. We see this relationship in the correlation plot that are negatively correlated with chocolate. 

```{r}
ggplot(pca$rotation) +
  aes(PC1, 
      reorder(rownames(pca$rotation), PC1)) +
  geom_col()
```


```{r}
plot(pca$x[,1:2])
```


```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```
```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
#library(plotly)
#ggplotly(p)
```


> Q25. Based on your exploratory analysis, correlation findings, and PCA results, what combination of characteristics appears to make a “winning” candy? How do these different analyses (visualization, correlation, PCA) support or complement each other in reaching this conclusion?


Winning are usually chocolate and often containing nuts/caramel and are usually not fruity or hard.This was supported by three matrices exploratory plots which showed the overall distribution of popularity by win percentage, correlation matrix which elucidated how candy characteristics are related (e.g. price point and chocolate are positively correlated), and PCA which compressed and clustered the data into less dimensions and showed that nutty chocolates have higher winpercents. 

