---
title: "Class 12"
author: "Jacob Hizon A17776679"
format: pdf
---

## Section 1. Proportion of G/G in a population 

Dow loaded a CSV file from Ensemble < https://www.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39894595-39895595;v=rs8067378;vdb=variation;vf=959672880#373531_tablePanel

Here we read this CSV file 

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
table(mxl$Genotype..forward.strand.)/nrow(mxl) *100
```



Now let's look at a different population. I selected the GBR. 
```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
```


Find proportion of G|G 

```{r}
round(table(gbr$Genotype..forward.strand.) /nrow(gbr) *100, 2)
```

This variant is associated with higher G|G frequencies.
Let's investigate this further


## Section 4: Population Scale Analysis [HOMEWORK]

> Q. 13 One sample is obviously not enough to know what is happening in a population. You are
interested in assessing genetic differences on a population scale. So, you processed about
~230 samples and did the normalization on a genome level. Now, you want to find whether
there is any association of the 4 asthma-associated SNPs (rs8067378…) on ORMDL3
expression

How many samples do we have?

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```

```{r}
table(expr$geno)
```

```{r}
library(ggplot2)
```
Let's make a boxplot

```{r}
ggplot(expr) + aes(geno, exp, fill=geno) +
  geom_boxplot(notch = TRUE)
```

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3? Hint: An example boxplot is provided overleaf – yours does not need
to be as polished as this one.


The boxplot above indicates that ORMDL3 expression differs by rs8067378 genotype. People with the A/A genotype have the highest expression, A/G individuals show intermediate expression, and G/G individuals have the lowest expression. Ultimately Since G/G had the lowest expression, the G allele is associated with the lowest ORMDL3 expression. 


