---
title: "Class 5: Data Viz with ggplot"
author: "Jacob Hizon (PID: A17776679)"
format: html
---

## Background

There are lot's of ways to make figures in R. These include so-called "base R" graphics (e.g. `plot()`) and tones of add-on packages like **ggplot2**

For example here we make the same plot with both:

```{r}
head(cars)
```

```{r}
plot(cars)
```
First I need to install the package with the command `install.packages()`.

> **N.B.** We never run an install cmd in a quarto code chunk or we will end up re-installing packages many many times - which is not what we want!


Every time we want to use one of these "add-on" packages we need to load it up in R with the `library()` function: 

```{r}
library(ggplot2)
```
```{r}
ggplot(cars)
```

Every ggplot needs at least 3 things:

- The **data**, the stuff you want plotted 
- The **aes**thitics, how the data map to the plot 
- The **geom**etry, the type of plot 

```{r}
ggplot(cars) + 
  aes(x = speed, y = dist) +
  geom_point()
```

Add a line to better show relationship between speed and dist 

```{r}
p <- ggplot(cars) + 
  aes(x = speed, y = dist) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) +
  labs(title = "Stopping distance of old cars",
       subtitle = "Data from the 'cars' object",
       x = "Speed (mph)",
       y = "Distance (ft)")
```

```{r}
p
```

```{r}
p + theme_bw()
```





## Gene expression plot

We can read the input data from the class website 

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

A first version plot 

```{r}
ggplot(genes) + 
  aes(Condition1, Condition2) + 
  geom_point()
```


```{r}
table( genes$State )
```


Version 2 let's color by `State` so we can see the up and down significant genes compared to all the "unchanging" genes 

```{r}
ggplot(genes) + 
  aes(Condition1, Condition2, col=State) + 
  geom_point()
```



Version 3 plot, let's modify the default colors to something we like 

```{r}
ggplot(genes) + 
  aes(Condition1, Condition2, col=State) + 
  geom_point() + 
  scale_color_manual(
    values=c("orange", "gray", "purple")
  )
```


```{r}
ggplot(genes) + 
  aes(Condition1, Condition2, col=State) + 
  geom_point() + 
  scale_color_manual(
    values=c("orange", "gray", "purple")) + 
      labs(x="Control (no drugs)", y="Drugs",
           title = "Gene Expression Changes upon GLP-1") +
      theme_minimal()
  
```

## Going Further

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```
```{r}
head(gapminder, 10)
```
```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col=continent) +
  geom_point(alpha = 0.3)

```

Let's "facet" (i.e. make a seperate plot) by continent rather than big hot mess above.

```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col=continent) +
  geom_point(alpha = 0.3) + 
  facet_wrap(~continent)

```
