---
title: "Pertussis and the CMI-PB project (Class 18)"
author: "Jacob Hizon A17776679"
format: pdf
---

# Background

Pertussis (whooping cough) is a common lung infection caused by the bacteria B. pertussis. It can infect anyone but it is most deadly for infants (under 1 year of age). 

## CDC tracking data
The CDC tracks the number of Pertussis cases: 

```{r, echo=FALSE}
cdc <- data.frame(
  year = c(
    1922L,
    1923L,
    1924L,
    1925L,
    1926L,
    1927L,
    1928L,
    1929L,
    1930L,
    1931L,
    1932L,
    1933L,
    1934L,
    1935L,
    1936L,
    1937L,
    1938L,
    1939L,
    1940L,
    1941L,
    1942L,
    1943L,
    1944L,
    1945L,
    1946L,
    1947L,
    1948L,
    1949L,
    1950L,
    1951L,
    1952L,
    1953L,
    1954L,
    1955L,
    1956L,
    1957L,
    1958L,
    1959L,
    1960L,
    1961L,
    1962L,
    1963L,
    1964L,
    1965L,
    1966L,
    1967L,
    1968L,
    1969L,
    1970L,
    1971L,
    1972L,
    1973L,
    1974L,
    1975L,
    1976L,
    1977L,
    1978L,
    1979L,
    1980L,
    1981L,
    1982L,
    1983L,
    1984L,
    1985L,
    1986L,
    1987L,
    1988L,
    1989L,
    1990L,
    1991L,
    1992L,
    1993L,
    1994L,
    1995L,
    1996L,
    1997L,
    1998L,
    1999L,
    2000L,
    2001L,
    2002L,
    2003L,
    2004L,
    2005L,
    2006L,
    2007L,
    2008L,
    2009L,
    2010L,
    2011L,
    2012L,
    2013L,
    2014L,
    2015L,
    2016L,
    2017L,
    2018L,
    2019L,
    2020L,
    2021L,
    2022L,
    2023L,
    2024L,
    2025L
  ),
  cases = c(
    107473,
    164191,
    165418,
    152003,
    202210,
    181411,
    161799,
    197371,
    166914,
    172559,
    215343,
    179135,
    265269,
    180518,
    147237,
    214652,
    227319,
    103188,
    183866,
    222202,
    191383,
    191890,
    109873,
    133792,
    109860,
    156517,
    74715,
    69479,
    120718,
    68687,
    45030,
    37129,
    60886,
    62786,
    31732,
    28295,
    32148,
    40005,
    14809,
    11468,
    17749,
    17135,
    13005,
    6799,
    7717,
    9718,
    4810,
    3285,
    4249,
    3036,
    3287,
    1759,
    2402,
    1738,
    1010,
    2177,
    2063,
    1623,
    1730,
    1248,
    1895,
    2463,
    2276,
    3589,
    4195,
    2823,
    3450,
    4157,
    4570,
    2719,
    4083,
    6586,
    4617,
    5137,
    7796,
    6564,
    7405,
    7298,
    7867,
    7580,
    9771,
    11647,
    25827,
    25616,
    15632,
    10454,
    13278,
    16858,
    27550,
    18719,
    48277,
    28639,
    32971,
    20762,
    17972,
    18975,
    15609,
    18617,
    6124,
    2116,
    3044,
    7063,
    22538,
    21996
  )
)
```

```{r}
cdc
```

> Q.I want a plot of year on the x vs cases on the y. 

```{r}
library(ggplot2)

ggplot(cdc) + 
  aes(year, cases) +
  geom_point() +
  geom_line()
  
```


> Q. Add annotation lines for the major milestones of wP vaccination roll-out (1946) and the switch to the aP vaccine (1996). What do you notice?

We see that there is an increase past 2020. 


```{r}
ggplot(cdc) + 
  aes(year, cases) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 1946, col="blue", lty =2) +
  geom_vline(xintercept = 1996, col="red", lty=2) + 
  geom_vline(xintercept = 2020, col="gray", lty=2)
  
```

> Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

I notice a sharp decline in cases from the introduction of the wP vaccine to the introduction of the 1996 aP vaccine. In 1996 when the new aP vaccine was introduced, we see cases nearly at zero. 


> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

After the introduction of the aP vaccine, there is a steady increase after 2000 with a sharp peak around ~2010. However, there is again a steady increase after 2020 due to vaccine hesitancy, antibiotic overuse and AMR. This can also suggest that there is a waning efficacy of the aP vaccine is lower than that of the wP vaccine, thus requiring booster vaccinations for the aP. 

# Exploring CMI-PB data

The CMI-PB project link: < https://www.cmi-pb.org > 

The project's mission of CMI-PB is to provide the scientific community with a comprehensive, high-quality and freely accessible resource of Pertussis booster vaccination.

Basically, make available a large dataset on the immune response to Pertussis. They use a "booster" vaccination as a proxy for Pertussis infection. 

They make their data available as JSON format API. We can read this into R with the `read_json()` function from the **jasonlite** package: 

```{r}
library(jsonlite)
subject <- read_json("https://www.cmi-pb.org/api/v5_1/subject", 
                     simplifyVector = TRUE)
head(subject)
```

> Q. How many aP and wP individuals are there in this `subject` table? 

```{r}
table(subject$infancy_vac)
```

> Q. How many male/female?

```{r}
table(subject$biological_sex)
```


>Q. What is the bread down of biological sex and race?
>Q. Is this representative of the US population?

```{r}
table(subject$race, subject$biological_sex)
```


We can now read more tables from the CMI-PB database 

```{r}
specimen <- read_json("https://www.cmi-pb.org/api/v5_1/specimen", 
                      simplifyVector = TRUE)
head(specimen)

ab_titer <- read_json("https://www.cmi-pb.org/api/v5_1/plasma_ab_titer", 
                      simplifyVector = TRUE)

```

```{r}
head(ab_titer)
```

To make sense of all this data, we need to "join" (a.k.a "merge" or "link") all these tables together. Only then will you know that a given antibody measurement (from the `ab_titer` table) was collected on a certain data from a certain wP or aP subject (from the `subject` table). 

We can use **dplyr** and the `*_join()` family of functions to do this. 


```{r}
library(dplyr)
meta <- inner_join(subject, specimen)
head(meta)
```
Let's do one more `inner_join()` to join the `ab_titer` with all our `meta` data.

```{r}
abdata <- inner_join(ab_titer, meta)
head(abdata)

```

> Q. How many different Ab "isotypes" values are in this dataset 

```{r}
table(abdata$isotype)
```

> Q 12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?


```{r}
table(abdata$dataset)
```

We see that there is there is double the amount of values in the 2023 dataset then there are in the 2022 dataset. 


> Q. How many different "antigen" values are measured 

```{r}
table(abdata$antigen)
```

Let's focus on IgG isotype 

Note: there are two ways to write the pipe command; |> or %>%

```{r}
igg <-abdata |>
      filter(isotype=="IgG")

head(igg)
```

Make a plot of `MFI_normalised` values for all `antigen` values. 

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot()
```


The antigens "PT", "FIM2/3" and "FHA" appear to have the widest range because the aP vaccines is composed of these antibodies. 

> Q. IS there a difference for these responses between aP and wP individuals. 

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac) +
  geom_boxplot()
```


```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() +
  facet_wrap(~infancy_vac)
```

> Q. 13 Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() +
  xlim(0,75) +
  facet_wrap(vars(visit), nrow=2)
```
> Q. 14 What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?
 
 The antigens that show the most obvious differences in IgG antibody levels over time are FIM2/3 and FHA. FIM2/3, which  are the antigen components the whooping cough bacterium, recognition increases over time due to  a boosted antigen-specific IgG response after vaccination. 
 
> Q. Is there a difference with time (i.e. before booster shot vs after booster shot)

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac) +
  geom_boxplot() +
  facet_wrap(~visit)
```

Let's look specifically at 2021 IgA in wP and aP 

```{r}
## filter to 2021 dataset, IgG and PT only 
abdata.21 <- abdata %>% filter(dataset == "2021_dataset")

abdata.21 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
     geom_smooth(
    aes(group = infancy_vac, col = infancy_vac),
    method = "loess",  # or method = "lm"
    se = FALSE,
    linewidth = 1.2
  )+
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```
ADD A TREND LINE FOR WP and AP seperately 
