---
title: "Class 8 Mini-Project"
author: "Jacob Hizon A17776679"
format: pdf
toc: true
---

## Background 

In today's class we will apply the methods and techniques clustering and PCA to help make sense of a real world breast cancer study FNA biopsy data set


## Data import 

We start by importing our data. It is a CSV file so we will use the `read.csv()` function. 

```{r}
wisc.df <- read.csv("WisconsinCancer.csv", row.names = 1)
```

```{r}
head(wisc.df)
```

Make sure to remove `diagnosis` column - I don't want to use this for my machine learning mdoels. We will use it later to compare our results to the expert diagnosis. 

```{r}
wisc.data <- wisc.df[,-1]
diagnosis <- wisc.df$diagnosis
```

> Q1. How many observations are in this dataset?

```{r}
nrow(wisc.df)
```

> Q2. How many of the observations have a malignant diagnosis?

```{r}
sum(wisc.df$diagnosis == "M")
```


```{r}
table(wisc.df$diagnosis)
```


> Q3. How many variables/features in the data are suffixed with _mean?

```{r}
#colnames(wisc.data)
length(grep("_mean", colnames(wisc.data)))
```


## Principal Component Analysis

The main function is `prcomp()` and we want to make sure we set the optional argument `scale = TRUE`:

```{r}
# Check column means and standard deviations
colMeans(wisc.data)

apply(wisc.data,2,sd)
```

```{r}
# Perform PCA on wisc.data by completing the following code
wisc.pr <- prcomp(wisc.data, scale = TRUE )
```


```{r}
# Look at summary of results
summary(wisc.pr)
```



> Q4. From your results, what proportion of the original variance is captured by the first principal component (PC1)?

44.27%

> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

3 PCs are required.

> Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

7 PCs are required.


Our main PCA "score plot" or "PC plot" of results:

```{r}
library(ggplot2)
```

```{r}
ggplot(wisc.pr$x)+
  aes(PC1, PC2, col=diagnosis) +
  geom_point()
```

> Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

The clear seperation stands out to me the most. It's easy to understand because there is clear seperation between the two clusters. Malignent and benign diagnoses are clearly seperated.


```{r}
# Repeat for components 1 and 3
ggplot(wisc.pr$x) +
  aes(PC1, PC3, col=diagnosis) +
  geom_point()
```



> Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

In PC1 and PC3 there is more overlap between the two clusters. This makes sense because since we're using PC3, it captures less varience between the two. 


> Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC. Are there any features with larger contributions than this one?

The loading of concave.points_mean on the first principal component is -0.26085.
Features such as area_mean have higher absolute loadings thus they contribute more to PC1 than concave.points_mean. 


```{r}
wisc.pr$rotation["concave.points_mean", 1]
```



```{r}
# Calculate variance of each component
pr.var <- wisc.pr$sdev^2
head(pr.var)
```



```{r}
# Variance explained by each principal component: pve
pve <- pr.var / sum(pr.var)

# Plot variance explained for each principal component
plot(c(1,pve), xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```

```{r}
# Alternative scree plot of the same data, note data driven y-axis
barplot(pve, ylab = "Percent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```

```{r}
## ggplot based graph
#install.packages("factoextra")
library(factoextra)
fviz_eig(wisc.pr, addlabels = TRUE)
```





## Hierarchical clustering

First scale the data (with the `scale()` function) then c

```{r}
# Scale the wisc.data data using the "scale()" function
data.scaled <- scale(wisc.data)
```

```{r}
data.dist <- dist(data.scaled)
```
```{r}
wisc.hclust <- hclust(data.dist, "complete")
```

```{r}
plot(wisc.hclust)
abline(h = 20, col="red", lty=2)
```

You can also use the `cutree()` function with a argument `k=4`
> Q10. Using the `plot()` and `abline()` functions, what is the height at which the clustering model has 4 clusters?

The height at which the clustering model has 4 clusters is approximately 20.

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k =4)
```
```{r}
table(wisc.hclust.clusters)
```

> Q11. OPTIONAL: Can you find a better cluster vs diagnoses match by cutting into a different number of clusters between 2 and 6? How do you judge the quality of your result in each case?

k =2 provides the  the best match between clusters and diagnoses in terms of interpretability. It becomes harder to intepret the results as you increase k. 

> Q12. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

My favorite method for clustering the data.dist dataset is “ward.D2”.

This is because it produces compact clusters that capture the total within-cluster variance, which is important for separating groups like benign vs. malignant diagnoses.

## Combining methods



Here we will take our PCA results and use those as inputs for clustering. In other words our `wisc.pr$x` scores that we plotted above (the main input of PCA - how the data lie on our new principal component acix/variables) and use a subset of the PCs as input for `hclust()`. 

```{r}
pc.dist <- dist(wisc.pr$x[,1:3])
wisc.pr.hclust <- hclust(pc.dist, method = "ward.D2")
plot(wisc.pr.hclust)
```


Cut the dendrogram into two main groups/clusters


```{r}
grps <- cutree(wisc.pr.hclust, k=2)
table(grps)
```


I want to know how clustering in `grps` with values 1 or 2 correspond the expert `diagnosis`
```{r}
table(grps, diagnosis)
```
My clustering **groups 1** are mostly "M" diagnosis (179) and my clustering **group 2** are mostly "B" diagnosis.

24 FP
179 TP
333 TN
33 FN 

```{r}
ggplot(wisc.pr$x) +
  aes(PC1, PC2) +
  geom_point(col=grps)
```
```{r}
# Use the distance along the first 7 PCs for clustering
wisc.pr.hclust <- hclust(dist(wisc.pr$x[, 1:7]), method = "ward.D2")
wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k = 2)
table(wisc.pr.hclust.clusters, diagnosis)
```
> Q13. How well does the newly created hclust model with two clusters separate out the two “M” and “B” diagnoses?

Overall, The PCA-based hierarchical clustering model does a good job of separating the two diagnoses. Cluster 1 mostly contains malignant samples (188), and Cluster 2 mostly contains benignsamples (329). There is some slight misclassification nonetheless. 



> Q14. How well do the hierarchical clustering models you created in the previous sections (i.e. without first doing PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.hclust.clusters and wisc.pr.hclust.clusters) with the vector containing the actual diagnoses.

The hierarchical clustering model without PCA created 4 clusters clusters 1 and 3 aligning the most accurately.
Cluster 1 had mostly malignant with 165, and cluster 3 had mostly benign with 343.
However, clusters 2 and 4 were small and out of place, reducing clarity.
This model overall is less clean as there are two other groups that seem out of place (2 and 4). 

```{r}
table(wisc.hclust.clusters, diagnosis)
```

## Sensitivity/Specificity


> Q15. OPTIONAL: Which of your analysis procedures resulted in a clustering model with the best specificity? How about sensitivity?


Our PCA outputs provided the best sensitivity while our output without PCA provided the better specificity. 


## 7. Prediction


```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

```{r}
plot(wisc.pr$x[,1:2], col=grps)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

> Q16. Which of these new patients should we prioritize for follow up based on your results?

We should prioritize patient's in cluster 2 because from previous PCA analysis we see that most of cluster 2 do indeed have true malignancies. 